EuPPollNet: A European database of plant-pollinator networks

### 2024-06-14 Manuscript/EuPPollNet_word.qmd
updating authors

**Motivation:** Pollinators play a crucial role in maintaining Earth’s terrestrial biodiversity and human food production by mediating sexual reproduction for most flowering plants. Indeed, the intricate network of interactions formed by plants and pollinators constitutes the backbone of plant-pollinator community stability and functioning. However, rapid human-induced environmental changes are compromising its long-term persistence. One of the major challenges for pollinator conservation is the lack of robust generalisable data capturing how plant-pollinator communities are structured across space and time. Here, we present the EuPPollNet database, a fully open and reproducible European-level database containing harmonized taxonomic data on plant-pollinator interactions referenced in both space and time. This database offers an open workflow that allows researchers to track data-curation decisions and edit them according to their preferences, while also providing other ecological variables of interest. Furthermore, this work provides an in-depth assessment of the taxonomic and sampling coverage of the database at the European level, complemented by analyses of key structural properties in plant-pollinator networks. We hope this database can help researchers to: 1) identify taxonomic, ecological, and geographical gaps of knowledge on plant-pollinator interactions; and 2) explore the impacts of global change on plant-pollinator networks to guide future conservation planning for both plant and pollinator species.

**Main Types of Variables Included:** EuPPollNet contains `r prettyNum(interaction_number, big.mark=",")` interactions between plants and pollinators from `r prettyNum(n_sites, big.mark=",")` distinct locations (i.e., plant-pollinator networks), which belong to `r study_number` different studies distributed across `r country_number` European countries. In addition, information about sampling methodology, habitat type, bio-climatic region, and further taxonomic rank information for both plant and pollinator species are also provided.

@@

**Time period and grain:** Species interaction data was recorded between `r min_year` and `r max_year`. All records are time-referenced and most of the studies documented interactions in a single flowering season (`r format(round(one_year, 2), nsmall = 2)`%).

**Major taxa and level of measurement:** The database contains interaction data at the species level for `r format(round(species_level_int, 2), nsmall = 2)`% of the records, including a total of `r prettyNum(plant_spp_number, big.mark=",")` plant and `r prettyNum(pollinator_spp_number, big.mark=",")` pollinator species. The database covers `r format(round(plant_coverage, 2), nsmall = 2)`% of the European species of flowering plants, `r format(round(bee_coverage, 2), nsmall = 2)`% of bees, `r format(round(lepidoptera_coverage, 2), nsmall = 2)`% of butterflies, and `r format(round(syrphid_coverage, 2), nsmall = 2)`% of syrphids species at the European level.

**Software format:** The database was built with R software and is stored as ".rds" and ".csv" formats. The construction of the database is fully reproducible and can be accessed at the following PERMANENT LINKS.

@@

The interaction between plants and pollinators is one of the most well-documented mutualisms on Earth. Plant-pollinator interactions involve a great diversity of species, largely attributed to their coevolutionary history [@ollerton2017], and are critically important for terrestrial biodiversity and economic productivity. The synergistic effects of climate change with other global change pressures are threatening worldwide biodiversity [@sala2000; @bellard2014], including plant and pollinator species as well as their interactions [@goulson2015; @settele2016; @eichenberg2021]. Under this scenario, the increasing availability of biodiversity data plays a major role in our ecological understanding of species status, trends, and conservation [@heberling2021; @zattara2021]. However, our knowledge of plant and pollinator species and their network of interactions still exhibits major temporal, spatial and taxonomic biases [@archer2014; @troia2016; @poisot2021; @marshall2024], limiting our ability to effectively protect their biodiversity.

Europe is one of the continents with a larger amount of available biodiversity data [@proencca2017], yet still exhibits major gaps [@wetzel2018; @bennett2018]. While species checklists need to be treated carefully, especially at a macro-ecological scale [@grenie2023], the growing number of European plant and pollinator checklists along with occurrence data, is setting a foundation for the conservation of its flora and their pollinators. However, species richness is just one component of biodiversity and documenting the interaction between plants and pollinators is essential for understanding the fate of the species within an ecosystem [@jordano2016]. Numerous works have studied plant-pollinator interactions in the last decades, originating thousands of plant-pollinator interaction networks worldwide. Several initiatives have tried to integrate plant-pollinator interaction data into databases such as *Mangal* [@poisot2016] or *GloBI* [@poelen2014], resulting in numerous large scale comparative analyses that have enhanced our understanding of the ecology of plants and pollinators [e.g., European wild bee data trends\; @marshall2024]. Despite all these resources, Europe lacks accessible harmonized plant-pollinator interaction data that allow researchers to evaluate the state of the art of plant-pollinator interactions at European level, which will guide research efforts, conservation planning and will set a foundation for future global change research.

The interactions between different plant and pollinator species within a community form complex networks. Macro-ecological analyses of the topology of these networks have revealed common properties across them, such as truncated power-law degree distributions [@jordano1987] or nestedness [@bascompte2003]. In addition, this type of large scale analyses can help understanding landscape level processes that cannot be explored at the community level, such as ecological patterns across geographic regions [@olesen2002; @traveset2016] or environmental gradients [@ramos2010; @rech2016; @saunders2023]. Although macro-ecological approaches make significant contributions to knowledge, they tend to be rare, and are strongly influenced by the spatio-temporal availability and nature of the data [@burkle2011; @trojelsgaard2013]. For instance, plant-pollinator studies tend to differ in sampling effort and methodology which can impact the structure of the resulting plant-pollinator networks [@gibson2011; @jordano2016]. Further, despite little sampling effort could capture most of the relevant functional species [@hegland2010], most plant-pollinator networks have unobserved interactions because undersampling [@olesen2011; @chacoff2012]. Thus, the lack of strong spatio-temporal coverage, along with the current intrinsic limitations of sampling plant-pollinator networks, highlight the need to keep gathering and integrating informative species interaction data to properly unravel the different ecological processes that shape plant-pollinator interactions at large scales.

Here, we present the EuPPollNet database, an acronym derived from European plant-pollinator networks, which contains harmonized information on interaction data of plants and pollinators at European level. The primary focus of the pollinator taxonomic groups is on the main orders of entomofauna that visit and pollinate flowering plants in Europe. These include insect species from the orders Hymenoptera, Diptera, Lepidoptera, and Coleoptera [@willemstein1987; @potts2015], which comprise almost the totality of recorded interactions in the EuPPollNet database (`r format(round(main_orders_percent, 2), nsmall = 2)`%). To understand the scope of the database, we examine the taxonomic and sampling coverage of the different plant and pollinator species at European level with the help of the most up-to-date species checklists and rarefaction analyses. In addition, we investigate how two key structural metrics of plant-pollinator networks, such as connectance and nestedness, change across the latitudinal range of studies and bioclimatic regions. Finally, we also examine whether plant-pollinator networks fulfill the expectation of being more nested than null expectations. EuPPollNet aims to cover a wide range of taxonomic groups and habitats, while also providing other variables of interest that allow better control for the ecological context and sampling methods. In addition, EuPPollNet offers a transparent and accessible workflow of its data management and species harmonization that allows it to be reused and keep building on it over time. This database provides a large number of community-level networks with curated and harmonized data, distinguishing it from other currently available resources that contain plant-pollinator interactions. We expect that this database will help to evaluate macro-ecological processes and current gaps of plant-pollinator interactions at European level.

@@

This database is the result of one of the working packages of the European project Safeguard. The EuPPollNet database includes published and unpublished studies compiled initially by a wide number of researchers and institutions within the European continent. Data acquisition followed a non-systematic approach. First, data was directly asked to members of the Safeguard project and then, the request was extended to data owners outside of the project. These other studies were identified by direct communication with other colleagues and by directly searching studies on Google Scholar of under-represented regions within the database. To keep high quality standards that will allow robust future ecological research, we only included studies meeting the following criteria: 1) studies that contained time- and geo-referenced records of plant-pollinator interactions; and 2) studies with phyto-centric plant-pollinator networks with quantitative visitation data.

@@

The database contains `r study_number` independent published and unpublished studies conducted during the time period `r min_year` - `r max_year` on `r country_number` different countries (**Figure 1a** and **Figure 1b**). The different studies differ in sampling effort and methodology, although most studies took place within a single flowering season (`r format(round(one_year, 2), nsmall = 2)`%), sampled a given location an average of `r format(round(average_sampling_days, 2), nsmall=2)` days, and documented interactions mostly by using `r  most_common_method[[1]]` as sampling method (`r format(round(most_common_method[[3]], 2), nsmall = 2)`%). The database includes a total of `r prettyNum(interaction_number, big.mark=",")` distinct interactions, considering interaction as the contact of a given pollinator to the reproductive structure of a particular plant. Most of the pollinator species in the database belong to the orders Hymenoptera and Diptera, each comprising approximately 1,000 species. However, the majority of plant-pollinator interactions are from Hymenoptera species (`r format(round(Hymenoptera_interactions, 2), nsmall = 2)`%; **Figure 1c**). Notably,  _Apis mellifera_ represents `r format(round(apis_percent, 2), nsmall = 2)`% of the total interaction records from the database and an average of `r format(round(apis_average_network, 2), nsmall = 2)`% of the total interactions per network.

```{r echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, out.width="110%", fig.cap= "\\textbf{Figure 1}. (\\textbf{a}) Approximate locations of the studies of the EuPPollNet database showing the total number of pollinator (i.e., orange heptagon) and plant (i.e., green circles) species per study. The sizes of these shapes are proportional to the respective species counts. For visualization purposes, we have focused only on the European region with studies and selected a single location per study. (\\textbf{b}) Number of studies by year within the database. (\\textbf{c}) Number of species and interactions across the 4 main pollinator orders at European level."}

@@

The EuPPollNet database is available in both .csv and .rds formats. The .csv file contains data in long format, while the .rds file stores the data in a list structure, with networks organized within each study. The file contains a total of `r nrow(my_tibble) + 1` columns that depict where and when the plant-pollinator interaction was observed (**Table 1**). To provide a standard unit of plant-pollinator interaction across studies, each row depicts a single interaction between a plant and a pollinator species. To build a plant-pollinator network matrix at a site level, authors only need to group interactions by plant and pollinator species, site and study. Information about habitat type and bioclimatic region are also provided in this file. In addition, the interaction dataset includes a column that describes the presence or absence of floral counts for each study. The flower count data is provided in a separate file given that one third of studies lack floral measurements and that the units differ greatly across studies. Finally, metadata at the study level is provided, including information about the authors, digital object identifier if available, sampling time and taxonomic coverage of the main pollinator groups for each study.

@@

@@

The completeness of the EuPPollNet database was evaluated by exploring the rarefied accumulation curves of plant and pollinator species and their interactions across the different sampling sites. In addition, an outstanding question in ecology is how many pollinators are required to pollinate flowering plants [@kleijn2015]. To that end, we also calculated the accumulation curve of pollinator species with increasing number plant species. The rarefied and extrapolated sampling curves were obtained with the help of the **iNEXT** package [@hsieh2016]. This was complemented with 100 random accumulation curves obtained programmatically to guide the visualization of the different rarefied curves.

All plant and pollinator species names were checked and standardized according to large scale taxonomic databases. To ensure reproducibility of the workflow, we have conducted this harmonization in R with **rgbif** [@chamberlain2022] as pivotal package to check for species names and retrieve further taxonomic information (i.e., phylum, order, family and genus) from the Global Biodiversity Information Facility (GBIF). The protocol for plants and pollinators is similar but slightly different given the availability of the different taxonomic resources. For transparency, we have included in the database the old species name, the new assigned name, and, if the name of the species is uncertain (e.g., species complex or species alike).

For plants: (i) we initially verified the exact matches against the GBIF species checklist; (ii) selected unmatched cases and fix orthographic errors; (iii) retrieved again taxonomic information for those unmatched cases, evaluated accuracy of fuzzing matching and programmatically fixed records that are still not found; (iv) finally, we used the World Flora Taxonomic Backbone [@govaerts2021\; WFO, July 7, 2022] as the ultimate filter for taxonomic information as we used it to calculate the plant taxonomic coverage of our database.

For pollinators: (i) we first created a checklist of species names for the most representative pollinator groups at the European level by combining the most up to date published checklists of bees and syrphids [@reverte2023] with the one of butterflies [@wiemers2018]; (ii) then, we compared pollinator species names against the checklist and recovered some unmatched cases with restrictive fuzzy matching by using **stringdist** package [@van2014]; (iii) we programmatically fixed unmatched records when necessary and retrieved the taxonomic information for all species from GBIF; (iv) we fixed the non-found cases in the GBIF checklist and made sure that all species names from bees, syrphids and butterflies were named according to their respective species checklists.

To assess the completeness of plant and pollinator species in the EuPPollNet database at European level, we used the aforementioned checklists for plants and pollinators. Specifically for plants, we refined the checklist to include only European flowering plants and excluded taxonomic groups not associated with biotic pollination. We did this by first excluding the families considered to have exclusively a wind pollination mode [see @culley2002], and then by filtering out the genera with wind or non-biotic pollination from families that exhibit both biotic and non-biotic pollination modes. Additionally, we manually included some exotic species and added unresolved species names that were not present in the accepted names of the checklist at the current version of usage. For pollinators, we compared only the taxonomic coverage of bees, syrphids and butterflies by using their species checklists at European level. The potential number of pollinator species at European level was estimated by adding the total number of species of bees, syrphids, and flies from the checklists, along with the extrapolated number of species from other insect pollinators. This extrapolated number was estimated by assuming that the sampling coverage of these other insect pollinators is equal to the average coverage across bees, syrphids, and butterflies (mean coverage = `r round(mean_coverage, 2)`; sd = `r round(sd_coverage,2)`).

Finally, to evaluate if the presence-absence of interaction records for bees and flowering plants follows a phylogenetic pattern within the database, we calculated its phylogenetic signal  at genus and family level, respectively. The phylogenetic signal was calculated by using the _phylosig_ function from the **phytools** package [@revell2012]. We extracted the phylogenetic information for bees from a genus level phylogeny [@hedtke2013] and processed it with the help of the packages **ape** [@paradis2019], **MCMCglmm** [@hadfield2010] and **phytools**; and for plants the phylogenetic tree was obtained from a species level plant phylogeny [@smith2018] with the help of **rtree** package [@li2023].

@@

The different sites per study were described with a habitat type by the authors. As these habitats are not standardized across studies, they were standardized with the additional help of land cover information and visual checks on current satellite imagery. For each georeferenced site, the land cover information was extracted from Corine Land Cover (CLC, version 2018) with the help of the Terra package [@hijmans2022]. Based on the habitat classification from the authors and the CLC classification, we created habitat categories that intend to summarize the diversity of habitats in the EuPPollNet database (see habitat type definition in supplementary material). These categories allow a quick comparison and understanding of the habitat types from the database. However, we advise authors to revise this classification if they intend to rely on this field for their analyses as this is a non-fully objective process. Moreover, Europe is characterized by a great variety of environmental conditions that harbor different biota. Thus, to allow authors to explore set of studies that share similar environmental conditions and species, we assigned to each site a biogeographical region. The biogeographical regions were downloaded from the European Environment Agency (version 2016) and were matched to the different sites with the help of a spatial joint from the **sf** package [@pebesma2018].

@@

To provide a general overview of the main aspects of this set of plant-pollinator networks, we evaluated network patterns of connectance and nestedness across the latitudinal range of studies and bioclimatic regions. We selected these two network metrics as they are commonly evaluated in plant-pollinator network studies and capture structural properties with a straightforward interpretation. However, these network metrics are not independent of sampling effort (e.g., number of species) and to allow comparisons across them, we implemented “standardised” versions of connectance and nestedness. As connectance is negatively associated with network size [@jordano1987], we evaluated how network connectance was associated with the number of species  (i.e., log of geometric mean of plants and pollinators) and extracted the residuals from this association (i.e., residual connectance) as a measurement of corrected connectance. This was done with the help of a Beta regression implemented with the package **betareg** [@cribari2010]. Second, we implemented a metric of nedstedness (i.e., NODFc) that allows the comparison across networks as it corrects by connectance and the number of species [@song2017]. This metric was calculated with the help of the **maxnodf** package [@hoeppke2021], which employs a computational eﬀicient approach to calculate it. Both residual connectance and NODFc were used as dependent variables to evaluate their association with latitude. In addition, we explored the association between connectance and nestedness with the number of species per network by conducting Kendall rank correlation coefficient in order to be able to compare the strength of this association across both network metrics.

Finally, to compare if networks are more or less nested than expected by chance, we employed the traditional z-score approach with the widely used nestedness metric (NODF) from @almeida2008, as this approach allow us to compare our results with previous published nestedness analyses in plant-pollinator networks and only compares each unique network against their randomized versions. We calculated 100 null models for each network with the help of the vaznull function from the package **bipartite** [@dormann2008] that implements the null model from @vazquez2005. These null networks have the same connectance and number of plant and pollinator species as the empirical ones, but different marginal totals. Both connectance and nestedness (NODF) were estimated for each network by using the function _networklevel_ from **bipartite**.

@@

The estimated sampling coverage of plant and pollinator species within the database across the different sampling sites is approximately `r round(poll_coverage)`% for both taxonomic groups. This suggests that the rarefied accumulation curves of both plant and pollinator species exhibit already a "quasi-asymptotic" growth of species richness by considering the current number of sampling sites or networks (**Figure 2a-2b**). The predicted observed species richness by doubling the sampling effort on the already sampled habitat types within the database will only increase pollinator richness by `r format(round(poll_increase3, 2), nsmall=2)`% and plant richness by `r format(round(plant_increase3, 2), nsmall=2)`%. However, the sampling coverage of interactions is `r format(round(int_coverage, 2), nsmall=2)`%, and by doubling the sampling effort the predicted number of unique interactions recorded will have approximately a twofold increase (`r format(round(int_increase3, 2), nsmall=2)`%; **Figure 2c**). Similarly, to the plant and pollinator species accumulation curves across sampling sites, when we consider the accumulated pollinator richness across sampled plant species, this curve also shows a "quasi-asymptotic" growth with a sampling coverage value of `r format(round(plant_poll_coverage, 2), nsmall=2)`%. The predicted recorded pollinator species by doubling the number of plants sampled is expected to increase by `r format(round(plant_poll_increase3, 2), nsmall=2)`% (**Figure 2d**). In addition, when we explore how common or rare are plant and pollinator species across all sampling sites, we find that the majority of plant and pollinator species tend to be regionally specific, and only a minor portion of them is being shared across a broad range of sampling sites (**Figures 2e-2f**). For instance, most plants (`r format(round(plants_under_one_percent, 2), nsmall=2)`%) and pollinators (`r format(round(polls_under_one_percent, 2), nsmall=2)`%) are exclusively found in less than 1% of sampling sites; and the most common plant (*`r most_common_plant`*) and pollinator (*`r most_common_poll`*) species are only found in `r format(round(most_common_plant_percent, 2), nsmall=2)`% and `r format(round(most_common_poll_percent, 2), nsmall=2)`% of sampling sites, respectively.

>>>>

**Motivation:** Pollinators play a crucial role in maintaining Earth’s terrestrial biodiversity and human food production by mediating sexual reproduction for most flowering plants. Indeed, the network of interactions formed by plants and pollinators constitutes the backbone of plant-pollinator community stability and functioning. However, rapid human-induced environmental changes are compromising the long-term persistence of plant-pollinator interaction networks. One of the major challenges for pollinator conservation is the lack of robust generalisable data capturing how plant-pollinator communities are structured across space and time. Here, we present the EuPPollNet database, a fully open and reproducible European-level database containing harmonized taxonomic data on plant-pollinator interactions referenced in both space and time, along with other ecological variables of interest. This database offers an open workflow that allows researchers to track data-curation decisions and edit them according to their preferences. We present the taxonomic and sampling coverage of EuPPollNet, and summarize key structural properties in plant-pollinator networks. We hope EuPPollNet will stimulate future research that fills the taxonomic, ecological, and geographical data gaps on plant-pollinator interactions that we have identified. Further, the variation in the structure of the networks in EuPPollNet provides a strong basis for future studies aimed at quantifying drivers of plant-pollinator network change and guiding future conservation planning for plants and pollinators.

**Main Types of Variables Included:** EuPPollNet contains `r prettyNum(interaction_number, big.mark=",")` interactions between plants and pollinators from `r prettyNum(n_networks, big.mark=",")` distinct networks (i.e., distinct sampling event in space or time), which belong to `r study_number` different studies distributed across `r country_number` European countries. In addition, information about sampling methodology, habitat type, bio-climatic region, and further taxonomic rank information for both plant and pollinator species are also provided (i.e., family, order and phylum).

@@

**Time period and grain:** Species interaction data was collected between `r min_year` and `r max_year`. All records are time-referenced and most of the studies documented interactions within a single flowering season (`r format(round(one_year, 2), nsmall = 2)`%).

**Major taxa and level of measurement:** The database contains interaction data at the species level for `r format(round(species_level_int, 2), nsmall = 2)`% of the records, including a total of `r prettyNum(plant_spp_number, big.mark=",")` plant and `r prettyNum(pollinator_spp_number, big.mark=",")` pollinator species. The database covers `r format(round(plant_coverage, 2), nsmall = 2)`% of the European species of flowering plants, `r format(round(bee_coverage, 2), nsmall = 2)`% of bees, `r format(round(lepidoptera_coverage, 2), nsmall = 2)`% of butterflies, and `r format(round(syrphid_coverage, 2), nsmall = 2)`% of syrphid species at the European level.

**Software format:** The database was built with the R programming language and is stored as “.rds” and “.csv” formats. The construction of the database is fully reproducible and can be accessed at the following link: <https://github.com/JoseBSL/EuPPollNet.git>.

@@

Plant-pollinator interactions involve a great diversity of species, largely attributed to their coevolutionary history [@ollerton2017], and are critically important for terrestrial biodiversity and economic productivity. The synergistic effects of climate change with other global change pressures are threatening worldwide biodiversity [@sala2000; @bellard2014], including plant and pollinator species as well as their interactions [@goulson2015; @settele2016; @eichenberg2021]. Under this scenario, the increasing availability of biodiversity data plays a major role in our ecological understanding of species status, trends, and conservation [@heberling2021; @zattara2021]. However, our knowledge of plant and pollinator species and their network of interactions still exhibits major temporal, spatial and taxonomic biases [@archer2014; @troia2016; @poisot2021; @marshall2024], limiting our ability to effectively protect their biodiversity.

The interactions between different plant and pollinator species within a community form complex networks. Macro-ecological analyses of the topology of these networks have revealed common properties, such as truncated power-law degree distributions [@jordano1987] or nestedness [@bascompte2003]. Large-scale analyses across multiple studies can quantify patterns across geographic regions [@olesen2002; @traveset2016] or environmental gradients [@ramos2010; @rech2016; @saunders2023] that cannot be examined in a single study. Although macro-ecological approaches that use ecological interactions make significant contributions to knowledge [@windsor2023], such synthesis work must consider variation across studies in the spatio-temporal nature of the data [@burkle2011; @trojelsgaard2013]. For instance, plant-pollinator studies tend to differ in sampling effort and methodology which affect the structure of the resulting plant-pollinator networks [@gibson2011; @jordano2016; @schwarz2020]. Most plant-pollinator networks have unobserved interactions [@olesen2011; @chacoff2012], and thus research that attempts to synthesize across published studies must have access to raw data on interactions in order to statistically account for sampling effort and completeness. This emphasizes the importance of providing data in its rawest possible form in datasets that will be utilized for synthesis and macro-ecological studies.

Europe is one of the continents with a larger amount of available biodiversity data [@proencca2017], yet still exhibits major gaps [@wetzel2018; @bennett2018]. While species checklists need to be treated carefully, especially at a macro-ecological scale [@grenie2023], the growing number of European plant and pollinator checklists [@reverte2023], along with occurrence data [@zattara2021], is setting a foundation for the conservation of its flora and their pollinators. However, species richness is just one component of biodiversity and documenting the interaction between plants and pollinators is essential for understanding biodiversity change [@jordano2016]. Numerous works have studied plant-pollinator interactions in the last decades, originating thousands of plant-pollinator interaction networks worldwide. Several initiatives have tried to integrate plant-pollinator interaction data into databases such as *Mangal* [@poisot2016] or *GloBI* [@poelen2014], resulting in numerous large scale comparative analyses that have enhanced our understanding of the ecology of plants and pollinators [e.g., European wild bee data trends\; @marshall2024]. Despite all these resources, Europe lacks accessible harmonized plant-pollinator interaction data that allow researchers to evaluate plant-pollinator interactions at a European level, which will guide research efforts, conservation planning and will set a foundation for future global change research. For example, only over a dozen of European plant-pollinator networks are included in *Mangal*, while *GloBI* focuses on pairwise interactions disconnected from the community context. Assembling and curating the existing information on EU plant-pollinator networks will guide research efforts, conservation planning and will set a foundation for future global change research.

Here, we present the European Plant-Pollinator Networks database (EuPPollNet), which contains harmonized information on plant-pollinator interactions at the European level. The pollinator taxonomic groups include the main orders of entomofauna that visit and pollinate flowering plants in Europe. These comprise insect species from the orders Hymenoptera, Diptera, Lepidoptera, and Coleoptera [@willemstein1987; @potts2015], accounting for almost the totality of recorded interactions in EuPPollNet (`r format(round(main_orders_percent, 2), nsmall = 2)`%). To understand the scope of the database, we examined the taxonomic and sampling coverage of the different plant and pollinator species at the European level with the help of the most up-to-date species checklists and rarefaction analyses. The overall number of potential pollinators and flowering plants in Europe was estimated with the help of the different checklists and sampling coverage. Finally, for bees and plants, we evaluated if there is a phylogenetic signal in the presence-absence of interaction data.

We investigated how two key structural metrics of plant-pollinator networks, connectance and nestedness, change across latitudes and bioclimatic regions. Networks that are more nested than expected by chance exhibit greater redundancy in their ecological functions and more stability in the face of species loss [@thebault2010]. EuPPollNet contains one of the largest sets of plant-pollinator networks collated to date, providing a unique opportunity to examine the prevalence of nestedness, which has only been evaluated with a relatively small number of networks [@bascompte2003; @staniczenko2013; @payrato2019]. EuPPollNet aims to cover a wide range of taxonomic groups and habitats, while also providing other variables of interest that define the ecological context and sampling methods of the study. In addition, EuPPollNet offers a transparent and accessible workflow of its data management and species harmonization that allows the database to be reused and to expand over time. This database provides a large number of community-level networks with curated and harmonized data, distinguishing it from other currently available resources that contain plant-pollinator interactions. We expect that EuPPollNet can be used to evaluate macro-ecological processes in plant-pollinator networks, guide conservation planning, and set a baseline for global change research.

@@

The EuPPollNet database includes published and unpublished studies compiled initially by a wide number of researchers and institutions within the European continent. As this database is the result of one of the working packages of the European project Safeguard, first, data was directly asked from members of the Safeguard project. Second, the request was extended to data owners outside of the project. These other data owners were identified by direct communication with colleagues suggested by Safeguard members and by directly searching for studies on Google Scholar of under-represented regions within the database. To maintain high quality standards that will support robust future ecological research, we only included studies meeting the following criteria: 1) studies that contained time- and geo-referenced records of plant-pollinator interactions; and 2) studies with phyto-centric plant-pollinator networks with quantitative visitation data.

@@

The database contains `r study_number` independent published and unpublished studies conducted during the time period `r min_year` - `r max_year` in `r country_number` different countries (**Figure 1a** and **Figure 1b**). The studies differ in sampling effort and methodology, and thus documenting sampling methods and sampling effort is an important feature of EuPPollNet. Most studies took place within a single flowering season (`r format(round(one_year, 2), nsmall = 2)`%), sampled a given location an average of `r format(round(average_sampling_days, 2), nsmall=2)` days, and documented interactions mostly by using `r  most_common_method[[1]]` as the sampling method (`r format(round(most_common_method[[3]], 2), nsmall = 2)`%). The database includes a total of `r prettyNum(interaction_number, big.mark=",")` distinct interactions, considering interaction as the contact of a pollinator individual to the reproductive structure of a particular plant. Most of the pollinator species in the database belong to the orders Hymenoptera and Diptera, each comprising approximately 1,000 species in the database. However, the majority of plant-pollinator interactions are from Hymenoptera species (`r format(round(Hymenoptera_interactions, 2), nsmall = 2)`%; **Figure 1c**). Notably,  _Apis mellifera_ represents `r format(round(apis_percent, 2), nsmall = 2)`% of the total interaction records from the database and an average of `r format(round(apis_average_network, 2), nsmall = 2)`% of the total interactions per network.

```{r echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, out.width="110%", fig.cap= "\\textbf{Figure 1}. (\\textbf{a}) Locations of the studies in EuPPollNet showing the total number of pollinator (i.e., orange heptagon) and plant (i.e., green circles) species per study. The sizes of these shapes are proportional to the respective species counts. For visualization purposes, we have focused only on the European region with studies and selected a single location per study. (\\textbf{b}) Number of studies by year in EuPPollNet. (\\textbf{c}) Number of species and interactions across the four main pollinator orders in EuPPollNet."}

@@

The EuPPollNet database is available in both .csv and .rds formats. The files contain a total of `r nrow(my_tibble) + 1` columns that depict where and when the plant-pollinator interaction was observed (**Table 1**). Each row depicts a single interaction between a plant and a pollinator species. To build a plant-pollinator network matrix within a single flowering season at a site level, users only need to group interactions by plant and pollinator species, site, study and year. Information about habitat type and bioclimatic region are also provided in this file. In addition, the interaction dataset includes a column that describes the presence or absence of floral counts for each study. The flower count data is provided in a separate file (.csv or .rds) that can be merged with the interaction data through the "Flower_data_merger" column. Two thirds of studies contain information on floral abundance, but the methods and units differ greatly across studies. Finally, metadata at the study level is provided, including information about the authors, digital object identifier if available, sampling time and taxonomic coverage of the main pollinator groups for each study.

@@

All plant and pollinator species names were checked and harmonized in R using **rgbif** [@chamberlain2022]. The protocol for plants and pollinators is similar but slightly different given the availability of the different taxonomic resources. For transparency, we have included in the database the original species name, the new assigned name, and, if the name of the species is uncertain (e.g., species complex or species alike). In addition, taxonomic information at family, order and phylum level was downloaded for each species.

For plants: (i) we initially verified the exact matches against the GBIF species checklist; (ii) we selected unmatched cases and fixed orthographic errors; (iii) we retrieved again taxonomic information for those unmatched cases, evaluated accuracy of fuzzing matching and programmatically fixed records that are still not found; (iv) finally, we used the World Flora Taxonomic Backbone [@govaerts2021\; WFO, July 7, 2022] as the ultimate filter for taxonomic information as we used it to calculate the plant taxonomic coverage of our database.

For pollinators: (i) we first created a checklist of species names for the most representative pollinator groups at the European level by combining the most up to date published checklists of bees and syrphids [@reverte2023], and butterflies [@wiemers2018]; (ii) we compared pollinator species names against the checklist and recovered some unmatched cases with restrictive fuzzy matching by using **stringdist** package [@van2014]; (iii) we fixed unmatched records when necessary and retrieved the taxonomic information for all species from GBIF; (iv) we fixed the non-found cases in the GBIF checklist and made sure that all species names from bees, syrphids and butterflies were named according to their respective species checklists. Coleoptera species names were only checked against the GBIF checklist.

To assess the completeness of plant and pollinator species in the EuPPollNet database at the European level, we used the aforementioned checklists for plants and pollinators. Specifically for plants, we refined the checklist to include only European flowering plants and excluded taxonomic groups not associated with biotic pollination. We did this by first excluding the families considered to have exclusively a wind pollination mode [see @culley2002], and then by filtering out the genera with wind or non-biotic pollination from families that exhibit both biotic and non-biotic pollination modes. Additionally, we manually included some exotic species and added unresolved species names that were not present in the accepted names of the checklist at the current version of usage. For pollinators, we compared only the taxonomic coverage of bees, syrphids and butterflies by using their species checklists at the European level. The number of pollinator species at the European level was estimated by first estimating coverage for the three taxa that have checklists (bees, syrphids and butterflies). The mean coverage across these three groups (mean coverage = `r round(mean_coverage, 2)`; sd = `r round(sd_coverage,2)`) was then used for the other taxa for which there is no checklist to extrapolate the number of pollinating species in Europe.

Finally, to evaluate if the presence-absence of interaction records for bees and flowering plants follows a phylogenetic pattern within the database, we calculated its phylogenetic signal  at genus and family level, respectively. The phylogenetic signal was calculated by using the _phylosig_ function from the **phytools** package [@revell2012]. We extracted the phylogenetic information for bees from a genus level phylogeny [@hedtke2013] and processed it using the packages **ape** [@paradis2019], **MCMCglmm** [@hadfield2010] and **phytools**. For plants, the phylogenetic tree was obtained from a species level plant phylogeny [@smith2018] with the help of the **rtree** package [@li2023].

@@

The completeness of the EuPPollNet database was evaluated by exploring the rarefied accumulation curves of plant and pollinator species and their interactions across the different networks. In addition, an outstanding question in ecology is how many pollinators are required to pollinate flowering plants [e.g., @kleijn2015 for crops]. To that end, we also calculated the accumulation curve of pollinator species with an increasing number of plant species. The rarefied and extrapolated sampling curves were obtained using the **iNEXT** package [@hsieh2016]. The different rarefied curves were complemented with 100 boostrapped accumulation curves.

@@

We describe the habitat type for each site using information from Corine Land Cover (CLC, version 2018) extracted using the Terra package [@hijmans2022], visual inspection of Google Earth imagery and the habitat classification from the authors. These different habitat categories (see definitions in supplementary text) allow a quick comparison and understanding of the habitat types from the database. Moreover, Europe is characterized by a great variety of environmental conditions that harbor different biota. Thus, to allow authors to explore the set of studies that share similar environmental conditions and species, we assigned a biogeographical region to each site. The biogeographical regions were downloaded from the European Environment Agency (version 2016) and were matched to the different sites using a spatial joint from the **sf** package [@pebesma2018].

@@

To provide a general overview of the structure of plant-pollinator networks in EuPPollNet, we quantified connectance and nestedness for each network, and examined how these network metrics change across latitude of studies and bioclimatic regions in Europe. We selected these two network metrics as they are commonly evaluated in plant-pollinator network studies and capture structural properties with a straightforward interpretation. We implemented “standardised” versions of connectance and nestedness to account for the effect of sampling effort on network metrics. As connectance is negatively associated with network size [@jordano1987], we evaluated how network connectance was associated with the number of species (i.e., log of geometric mean of plants and pollinators) and extracted the residuals from this association (i.e., residual connectance) as a measurement of corrected connectance. The relationship between residual connectance and species richness was investigated using a beta-regression implemented with the package **betareg** [@cribari2010]. We used NODFc to compare nestedness across networks, as it corrects by connectance and the number of species [@song2017]. This metric was calculated using the **maxnodf** package [@hoeppke2021]. Both residual connectance and NODFc were used as dependent variables to evaluate their association with latitude. In addition, we quantified the association between connectance and nestedness with the number of species per network using a Kendall rank correlation coefficient to compare the strength of associations between network structures and species richness across both network metrics.

Finally, to compare if networks are more or less nested than expected by chance, we employed the traditional z-score approach with the widely used nestedness metric (NODF) from @almeida2008. The z-score approach allows us to compare our results with previous published nestedness analyses in plant-pollinator networks and only compares each unique network against their randomized versions. We calculated 100 null models for each network with the help of the _vaznull_ function from the package **bipartite** [@dormann2008]. These null networks have the same connectance and number of plant and pollinator species as the empirical ones, but different marginal totals. Both connectance and nestedness (NODF) were estimated for each network by using the function _networklevel_ from **bipartite**.

@@
